How do changes in ex vary between countries?

### 2022-03-01 draft_paper.Rmd
add synthetic germany to series and nice labels

@@

@@

hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~code) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

@@

```{r}
change_in_ex_selected_countries <- 
  hmd_ex_selected_countries %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

@@

```{r}
change_in_ex_selected_countries %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(code~sex) + 
  geom_hline(yintercept = 0)

>>>>

### Synthetic Germany

- To maybe be moved to the appendix later

To allow UK national trends to be compared with a single German population, we attempted to produce a 'Synthetic German' population with data for East and West Germany for years prior to reunification. We estimated that this 'Synthetic Germany' could be produced by using a weighted average of 20% East Germany, and 80% West German life expectancy trends. More precise estimates can be produced, and the methods used to reach this conclusion are detailed in Appendix X.

@@

# Code for producing synthetic Germany

source(here("R", "make_synthetic_germany_functions.R"))

series_east_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_f0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_f0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "female", 
  ex = make_synthetic_population(series_east_f0, series_west_f0, p_east = 0.2)
)

###
series_east_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_f65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "female") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_f65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "female", 
  ex = make_synthetic_population(series_east_f65, series_west_f65, p_east = 0.2)
)
##

series_east_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_m0 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_m0 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 0,
  sex = "male", 
  ex = make_synthetic_population(series_east_m0, series_west_m0, p_east = 0.2)
)

###

series_east_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTE") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

series_west_m65 <- hmd_ex_selected_countries %>% 
  filter(code == "DEUTW") %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  filter(sex == "male") %>% 
  arrange(year) %>% 
  pull(ex)

tmp_df_m65 <- tibble(
  code = "DEUTSYNTH",
  year = 1980:2017,
  x = 65,
  sex = "male", 
  ex = make_synthetic_population(series_east_m65, series_west_m65, p_east = 0.2)
)

hmd_ex_selected_countries_with_synth <- bind_rows(
  hmd_ex_selected_countries,
  tmp_df_m0,
  tmp_df_m65,
  tmp_df_f0,
  tmp_df_f65
)

# Labels for codes

country_code_lookup <- 
  tribble(
    ~code, ~country,
    "DEUTNP", "Germany",
    "DEUTE", "East Germany",
    "DEUTW", "West Germany",
    "ESP", "Spain",
    "FRATNP", "France", 
    "ITA", "Italy",
    "GBRTENW", "England & Wales",
    "GBR_SCO", "Scotland",
    "DEUTSYNTH", "Synthetic Germany"
  )

@@

hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 0) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at birth",
    title = "Life expectancies at birth for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

@@

```{r, echo = FALSE, fig.cap = "Life Expectancy at age 65"}
hmd_ex_selected_countries_with_synth %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  filter(x == 65) %>% 
  ggplot(aes(x = year, y = ex, group = sex, colour = sex, linetype = sex)) + 
  geom_line() +
  facet_wrap(~country) +
  labs(
    x = "Year",
    y = "Life Expectancy at age 65",
    title = "Life expectancies at age 65 for selected nations",
    subtitle = "1980 to 2020 or latest available year",
    caption = "Source: Human Mortality Database"
  )

@@

```{r}
change_in_ex_selected_countries <- 
  hmd_ex_selected_countries_with_synth %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>% 
    ungroup()

@@

```{r}
change_in_ex_selected_countries %>% 
  filter(x == 0) %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "East Germany", "West Germany", "Germany", "Synthetic Germany", "Spain", "France", "Italy"))) %>% 
  filter(between(year, 1980, 2020)) %>% 
  ggplot(aes(x = year, y = delta_ex)) + 
  geom_point(alpha = 0.2) + 
  stat_smooth() + 
  facet_grid(country~sex) + 
  geom_hline(yintercept = 0) + 
  theme(strip.text.y = element_text(angle = 0)) + 
  labs(
    x = "Year",
    y = "Year-on-year change in life expectancy",
    title = "Annual change in life expectancy at birth, selected countries",
    subtitle = "Blue line: nonlinear smoother",
    caption =  "Source: Human Mortality Database. Synthetic Germany based on 20% East/80% West German population weighting"
  )
